Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 numerical-mhd-dedner.tex
Reformatting tex files so they're all standardized in terms of paragraph wrapping, etc. (no content changed, just formatting.)

To more clearly describe the Dedner divergence cleaning modifications, we write the equations of compressible inviscid hydrodynamics in the form of conservation laws as,
\begin{equation}
 \frac{\partial{U}}{\partial{t}} +
 \frac{\partial{F^x}}{\partial{x}} + \frac{\partial{F^y}}{\partial{y}} + \frac{\partial{F^z}}{\partial{z}}= 0, \label{hydro}
\end{equation}
The conserved variable $U$ is given by
\begin{equation}
 U = (\rho, \rho v_x, \rho v_y, \rho v_z, \rho E)^{T},
\end{equation} 
where $\rho$ is density, $v_i$ are the three components of velocity
for $i={x,y,z}$, $E=v^2/2 + e$ denotes the specific total energy and $e$ the
specific internal energy (note that in this section only, we use the specific
energy).

For the generalized Lagrange multiplier (GLM) formulation of the MHD
equations \citep{2002JCoPh.175..645D}, we consider these 
conserved variables
\begin{equation}
 U = (\rho, \rho v_x, \rho v_y, \rho v_z, \rho E+B^2/2, B_x, B_y, B_z, \psi)^{T},
\end{equation} 
where $B_i$ with $i={x,y,z}$ are the three components of magnetic
fields and $\psi$ is the additional scalar field introduced in the GLM
formulation for the divergence cleaning.  The fluxes then are
\begin{eqnarray}
 F^x &=& (\rho v_x, \rho v_x^2+p+B^2/2-B_x^2, \rho v_yv_x-B_yB_x, \cr
 && \rho v_zv_x-B_zB_x, \rho ({v^2\over2} + h)v_x+B^2v_x-B_xB\cdot v, \cr
&& \psi, v_xB_y-v_yB_x, -v_zB_x+v_xB_z, c_h^2B_x)^T,\\
 F^y &=& (\rho v_y, \rho v_xv_y-B_xB_y, \rho v_y^2+p+B^2/2-B_y^2, \cr
 && \rho v_zv_y-B_zB_y, \rho ({v^2\over2} + h)v_y+B^2v_y-B_yB\cdot v, \cr
 && v_yB_z-v_zB_y,\psi,-v_xB_y+v_yB_x, c_h^2B_y)^{T}, \\
 F^z &=& (\rho v_z, \rho v_xv_z-B_xB_z, \rho v_yv_z-B_yB_z, \rho v_z^2+p+B^2/2-B_z^2, \cr
 && \rho ({v^2\over2} + h)v_z+B^2v_z-B_zB\cdot v, \cr
    &&  -v_yB_z+v_zB_y, v_zB_x-v_xB_z,\psi, c_h^2B_z)^{T},
\end{eqnarray}
where $c_h$ is a constant controlling the propagation speed and
damping rate of $\div B$, and $h=e+p/\rho$ denotes the specific 
enthalpy.  All quantities are cell-centered.

>>>>

To more clearly describe the Dedner divergence cleaning modifications,
we write the equations of compressible inviscid hydrodynamics in the
form of conservation laws as,

\begin{equation}
 \frac{\partial{U}}{\partial{t}} +
 \frac{\partial{F^x}}{\partial{x}} + \frac{\partial{F^y}}{\partial{y}} + \frac{\partial{F^z}}{\partial{z}}= 0, \label{hydro}
\end{equation}

The conserved variable $U$ is given by

\begin{equation}
 U = (\rho, \rho v_x, \rho v_y, \rho v_z, \rho E)^{T},
\end{equation}

where $\rho$ is density, $v_i$ are the three components of velocity
for $i={x,y,z}$, $E=v^2/2 + e$ denotes the specific total energy and $e$ the
specific internal energy (note that in this section only, we use the specific
energy).

For the generalized Lagrange multiplier (GLM) formulation of the MHD
equations \citep{2002JCoPh.175..645D}, we consider these 
conserved variables

\begin{equation}
 U = (\rho, \rho v_x, \rho v_y, \rho v_z, \rho E+B^2/2, B_x, B_y, B_z, \psi)^{T},
\end{equation}

where $B_i$ with $i={x,y,z}$ are the three components of magnetic
fields and $\psi$ is the additional scalar field introduced in the GLM
formulation for the divergence cleaning.  The fluxes then are

\begin{eqnarray}
 F^x &=& (\rho v_x, \rho v_x^2+p+B^2/2-B_x^2, \rho v_yv_x-B_yB_x, \cr
 && \rho v_zv_x-B_zB_x, \rho ({v^2\over2} + h)v_x+B^2v_x-B_xB\cdot v, \cr
&& \psi, v_xB_y-v_yB_x, -v_zB_x+v_xB_z, c_h^2B_x)^T,\\
 F^y &=& (\rho v_y, \rho v_xv_y-B_xB_y, \rho v_y^2+p+B^2/2-B_y^2, \cr
 && \rho v_zv_y-B_zB_y, \rho ({v^2\over2} + h)v_y+B^2v_y-B_yB\cdot v, \cr
 && v_yB_z-v_zB_y,\psi,-v_xB_y+v_yB_x, c_h^2B_y)^{T}, \\
 F^z &=& (\rho v_z, \rho v_xv_z-B_xB_z, \rho v_yv_z-B_yB_z, \rho v_z^2+p+B^2/2-B_z^2, \cr
 && \rho ({v^2\over2} + h)v_z+B^2v_z-B_zB\cdot v, \cr
    &&  -v_yB_z+v_zB_y, v_zB_x-v_xB_z,\psi, c_h^2B_z)^{T},
\end{eqnarray}

where $c_h$ is a constant controlling the propagation speed and
damping rate of $\div B$, and $h=e+p/\rho$ denotes the specific
enthalpy.  All quantities are cell-centered.

### 2013-06-25 numerical-ppm.tex
Removed commented-out text from lots of files - basically vestigial junk.
Reformatting tex files so they're all standardized in terms of paragraph wrapping, etc. (no content changed, just formatting.)

%\subsubsection{The piecewise parabolic method in one dimension} % 2.2

@@

We note that the acceleration required in Equation~(\ref{eq:momentum_diff}) is actually the
acceleration felt by the entire zone and not just at the zone center.
Therefore, it is possible to find the mass-weighted average acceleration over the zone
by expanding the density and acceleration distributions and
retaining all terms up to second-order in $\Delta x$ \citep{1995CoPhC..89..149B}, although we find that
the potential is so slowly varying that this is unnecessary.
%
%\begin{equation}
%g_j^{n+1/2} = 
%       \frac{1}{2 a^{n+1/2} \delta x_j} \left[ 
%             \phi_{j+1}^{n+1/2} 
%           - \phi_{j-1}^{n+1/2} 
%           + \frac{1}{12} \left(    \phi^{n+1/2}_{j+1} 
%                                 - 2\phi^{n+1/2}_j 
%                                 + \phi^{n+1/2}_{j-1} \right) 
%                                   {\delta d_j \over d_j}
%       \right].
%       \label{eq:accel}
%\end{equation}
%Generally we find that the potential is so slowly varying that this change is unnecessary.

@@

It should be stressed that as long as the parameter ${\eta}_1$ is small enough
the dual energy method {\it will have no dynamical effect}.  We use
${\eta}_1 = 10^{-3}$, which is consistent with the truncation error of the
scheme for grid sizes that are typically used in our simulations.  We are now free to select the method by which the internal energy
field variable $e$ is updated so that it will not become contaminated with
errors advected by the total energy formulation but still give the
correct distribution in shocked regions. 
Since we are concerned with the advection of errors, the selection
criterion must look at each cell's local neighbourhood.
In one dimension, this is done with,
\begin{equation}
e = \cases{ (E - \vecv^2/2),& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) > {\eta}_2$,\cr
            e,& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) < {\eta}_2$.\cr}
    \label{eq:dualselect}
\end{equation}
Thus, ${\eta}_2$ determines when the synchronization (of $e$ with $E$) occurs.
%is a measure of how much error we will allow to be passed
%from neighboring cells.  
Too high a value may mask relatively weak shocks, 
while spurious heating (via contamination) may occur if it is set too low.
After some experimentation,
we have chosen ${\eta}_2 = 0.1$, a somewhat conservative value.
This scheme is optional and is generally only required in large-scale 
cosmological simulations where the gas cools due to the expansion of the universe
but large bulk flows develop due to the formation of structure.

>>>>

@@

We note that the acceleration required in
Equation~(\ref{eq:momentum_diff}) is actually the acceleration felt by
the entire zone and not just at the zone center.  Therefore, it is
possible to find the mass-weighted average acceleration over the zone
by expanding the density and acceleration distributions and retaining
all terms up to second-order in $\Delta x$
\citep{1995CoPhC..89..149B}, although we find that the potential is so
slowly varying that this is unnecessary.

@@

It should be stressed that as long as the parameter ${\eta}_1$ is
small enough the dual energy method {\it will have no dynamical
effect}.  We use ${\eta}_1 = 10^{-3}$, which is consistent with the
truncation error of the scheme for grid sizes that are typically used
in our simulations.  We are now free to select the method by which the
internal energy field variable $e$ is updated so that it will not
become contaminated with errors advected by the total energy
formulation but still give the correct distribution in shocked
regions.  Since we are concerned with the advection of errors, the
selection criterion must look at each cell's local neighbourhood.  In
one dimension, this is done with,

\begin{equation}
e = \cases{ (E - \vecv^2/2),& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) > {\eta}_2$,\cr
            e,& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) < {\eta}_2$.\cr}
    \label{eq:dualselect}
\end{equation}

Thus, ${\eta}_2$ determines when the synchronization (of $e$ with $E$)
occurs.  Too high a value may mask relatively weak shocks, while
spurious heating (via contamination) may occur if it is set too low.
After some experimentation, we have chosen ${\eta}_2 = 0.1$, a
somewhat conservative value.  This scheme is optional and is generally
only required in large-scale cosmological simulations where the gas
cools due to the expansion of the universe but large bulk flows
develop due to the formation of structure.

### 2013-06-25 numerical-timestep.tex
Reformatting tex files so they're all standardized in terms of paragraph wrapping, etc. (no content changed, just formatting.)

Equation~(\ref{eqn:dthydro}) is valid for one dimension, and is used
when the equations of hydrodynamics are being solved.  For 2 or 3
dimensions, it was shown by \cite{Godunov1959}  that using the
harmonic average of the timestep found along each of the coordinate
axes yields a maximum $\kappa_{\rm hydro} = 0.8$.  So letting $\Delta
t_x$, $\Delta t_y$, and $\Delta t_z$ be the analogues of
equation~(\ref{eqn:dthydro}) along the $x,y$ and $z$ axes, 
\begin{equation}
  \Delta t_{\rm hydro} = \min \left( \frac{\kappa_{\rm hydro}} {1/\Delta t_x
  +1/\Delta t_y + 1/\Delta t_z} \right)_L
\end{equation}

@@

Equations~(\ref{eqn:dtaccel}) and~(\ref{eqn:dtrad}) are supplementary to equation~(\ref{eqn:dthydro}) in that they
take into account the possibility of large accelerations due to either
gravity (equation~\ref{eqn:dtaccel}) or radiation pressure
(equation~\ref{eqn:dtrad}).  In equation~(\ref{eqn:dtaccel}), $\myvec{g}$ is the
gravitational acceleration in each cell on level l.  In
equation~(\ref{eqn:dtrad}), $\myvec{a}_{\rm rad}$ is the estimated
acceleration due to radiation pressure in each cell on level l,
defined as
\begin{equation}
\myvec{a}_{rad} = \frac{ \sum_i \frac{\dot{E_i}}{c} \hat{r_i} }{m_b} 
\end{equation}
where the sum calculates the energy deposited in a cell during the
previous timestep due to \textit{all photon packets} that crossed that
cell, with $\hat{r_i}$ being a unit vector that accounts for the
packet direction.

Equation~(\ref{eqn:dtcond}) is the stability condition for an explicit
solution to the equation of heat conduction.  In this expression,
$n_b$ is the baryon number density, 
$\kappa_{\rm sp}(T)$ is the Spitzer thermal conductivity, and f$_{\rm sp}$ is a
user-defined, dimensionless
conduction suppression factor whose value must be f$_{\rm sp} \leq 1$.  
%Both are defined in Section~\ref{sec.num.conductions}.  
$k_{\rm cond}$ is a dimensionless prefactor whose
value must be $0 < k_{\rm cond}  \leq 0.5$, and is exactly 0.5 for the
implemention in \enzo.
From a practical
perspective, it is useful to note that, unlike other timestep criteria
discussed above (which effectively scale as $\Delta x / \sqrt{T}$ with
the grid cell size $\Delta x$ and temperature $T$),
the timestep criterion due to thermal conduction scales as $\Delta x^2
/ T^{2.5}$, which can result in a rapid decrease in timestep in
regions of high resolution and/or temperature.

>>>>

Equation~(\ref{eqn:dthydro}) is valid for one dimension, and is used
when the equations of hydrodynamics are being solved.  For 2 or 3
dimensions, it was shown by \cite{Godunov1959} that using the harmonic
average of the timestep found along each of the coordinate axes yields
a maximum $\kappa_{\rm hydro} = 0.8$.  So letting $\Delta t_x$,
$\Delta t_y$, and $\Delta t_z$ be the analogues of
equation~(\ref{eqn:dthydro}) along the $x,y$ and $z$ axes,

\begin{equation}
  \Delta t_{\rm hydro} = \min \left( \frac{\kappa_{\rm hydro}} {1/\Delta t_x
  +1/\Delta t_y + 1/\Delta t_z} \right)_L
\end{equation}

@@

Equations~(\ref{eqn:dtaccel}) and~(\ref{eqn:dtrad}) are supplementary
to equation~(\ref{eqn:dthydro}) in that they take into account the
possibility of large accelerations due to either gravity
(equation~\ref{eqn:dtaccel}) or radiation pressure
(equation~\ref{eqn:dtrad}).  In equation~(\ref{eqn:dtaccel}),
$\myvec{g}$ is the gravitational acceleration in each cell on level l.
In equation~(\ref{eqn:dtrad}), $\myvec{a}_{\rm rad}$ is the estimated
acceleration due to radiation pressure in each cell on level l,
defined as

\begin{equation}
\myvec{a}_{rad} = \frac{ \sum_i \frac{\dot{E_i}}{c} \hat{r_i} }{m_b} 
\end{equation}

where the sum calculates the energy deposited in a cell during the
previous timestep due to \textit{all photon packets} that crossed that
cell, with $\hat{r_i}$ being a unit vector that accounts for the
packet direction.

Equation~(\ref{eqn:dtcond}) is the stability condition for an explicit
solution to the equation of heat conduction.  In this expression,
$n_b$ is the baryon number density, $\kappa_{\rm sp}(T)$ is the
Spitzer thermal conductivity, and f$_{\rm sp}$ is a user-defined,
dimensionless conduction suppression factor whose value must be
f$_{\rm sp} \leq 1$.  $k_{\rm cond}$ is a dimensionless prefactor
whose value must be $0 < k_{\rm cond} \leq 0.5$, and is exactly 0.5
for the implemention in \enzo.  From a practical perspective, it is
useful to note that, unlike other timestep criteria discussed above
(which effectively scale as $\Delta x / \sqrt{T}$ with the grid cell
size $\Delta x$ and temperature $T$), the timestep criterion due to
thermal conduction scales as $\Delta x^2 / T^{2.5}$, which can result
in a rapid decrease in timestep in regions of high resolution and/or
temperature.

### 2013-06-25 tests-orszagtang.tex
Reformatting tex files so they're all standardized in terms of paragraph wrapping, etc. (no content changed, just formatting.)

\subsubsection{MHD: Orszag-Tang Vortex}
\label{sec.tests.mhd}
Figure~\ref{fig.orszag} shows the Orszag-Tang vortex problem \citep{Orszag79}.  The left panel shows
the result using \enzo's constrained transport MHD method, while the right panel shows the result using
\enzo's implementation of Dedner MHD.
The Orszag-Tang vortex test is a classig MHD test problem, and shows that significant small scale structure can be generated in MHD
from large scale initial perturbations.  It is often used to compare the
effective resolution of different MHD schemes.  The test begins with uniform
density, $\rho_0=25/(36\pi)$, and pressure, $P_0=5/(12\pi)$ (as with
other tests in this section, in the absence of gravity or chemistry/cooling we use
dimensionless units).  There is a
single rotational mode in the velocity, and two in the magnetic field:
${\bf v}_0 = (-\sin(2\pi y) \hat{x},\, \sin(2\pi x) \hat{y})$,
${\bf B}_0 = (-\sin(2\pi y) \hat{x},\, \sin(4\pi x) \hat{y})$.
The simulation is evolved to $t=0.48$.  One can see that the structures
are accurately represented as compared to, for example,
\citet{Toth00}, and that the resolution of shocks is comparable in
both methods.

>>>>

\subsubsection{MHD: Orszag-Tang Vortex}
\label{sec.tests.mhd}

Figure~\ref{fig.orszag} shows the Orszag-Tang vortex problem
\citep{Orszag79}.  The left panel shows the result using \enzo's
constrained transport MHD method, while the right panel shows the
result using \enzo's implementation of Dedner MHD.  The Orszag-Tang
vortex test is a classig MHD test problem, and shows that significant
small scale structure can be generated in MHD from large scale initial
perturbations.  It is often used to compare the effective resolution
of different MHD schemes.  The test begins with uniform density,
$\rho_0=25/(36\pi)$, and pressure, $P_0=5/(12\pi)$ (as with other
tests in this section, in the absence of gravity or chemistry/cooling
we use dimensionless units).  There is a single rotational mode in the
velocity, and two in the magnetic field: ${\bf v}_0 = (-\sin(2\pi y)
\hat{x},\, \sin(2\pi x) \hat{y})$, ${\bf B}_0 = (-\sin(2\pi y)
\hat{x},\, \sin(4\pi x) \hat{y})$.  The simulation is evolved to
$t=0.48$.  One can see that the structures are accurately represented
as compared to, for example, \citet{Toth00}, and that the resolution
of shocks is comparable in both methods.

### 2013-06-25 tests-sodshocktube.tex
Reformatting tex files so they're all standardized in terms of paragraph wrapping, etc. (no content changed, just formatting.)

We begin the paper test suite with the classic one-dimensional Sod shock tube problem
\citep{Sod78}, which provides a good test of a hydrodynamical solver's
ability to resolve a clean Riemann problem with clear separation
between the three resultant waves.  These waves consist of a rarefaction fan, a
contact discontinuity, and a moderate shock.  The initial state is 
$\rho_{\rm L}, P_{\rm L} = 1.0, 1.0$ on the left of the boundary at 
$x=0.5$ and $\rho_{\rm R}, P_{\rm R} = 0.125, 0.1$ on the right.
All velocities are initially zero.  In
Figure~\ref{fig.sodshocktube}, we show the density solution at the
final time ($t=0.25$) for three of our hydro solvers -- the spatially third-order
PPM, as well as the two second-order \zeus\ and MUSCL schemes.  We use
100 cells across the domain, which is a relatively standard choice in
code method papers, and show solutions both with and without adaptive
mesh refinement.  Without AMR (top row in Figure \ref{fig.sodshocktube}), it is
clear that the PPM scheme produces by far the cleanest solution with
all wave families crisply reproduced (in particular, the contact
discontinuity and shock).  \zeus\ and MUSCL produce similar results,
with MUSCL doing a slightly better job on the rarefaction fan.  In Figure
\ref{fig.sodshocktube}, we also show the integrated absolute deviation from 
the exact solution, $||E_1|| = \sum_i \Delta x_i |F(x_i)-{\rm Exact}(x_i)|$.
%, as well as the maximum deviation,
%$||E_\infty|| = \mathrm{max_i}\left\{ \Delta x_i|F(x_i)-{\rm Exact}(x_i)| \right\}$.  
These numbers confirm the qualitative differences noted
previously.

>>>>

We begin the paper test suite with the classic one-dimensional Sod
shock tube problem \citep{Sod78}, which provides a good test of a
hydrodynamical solver's ability to resolve a clean Riemann problem
with clear separation between the three resultant waves.  These waves
consist of a rarefaction fan, a contact discontinuity, and a moderate
shock.  The initial state is $\rho_{\rm L}, P_{\rm L} = 1.0, 1.0$ on
the left of the boundary at $x=0.5$ and $\rho_{\rm R}, P_{\rm R} =
0.125, 0.1$ on the right.  All velocities are initially zero.  In
Figure~\ref{fig.sodshocktube}, we show the density solution at the
final time ($t=0.25$) for three of our hydro solvers -- the spatially
third-order PPM, as well as the two second-order \zeus\ and MUSCL
schemes.  We use 100 cells across the domain, which is a relatively
standard choice in code method papers, and show solutions both with
and without adaptive mesh refinement.  Without AMR (top row in Figure
\ref{fig.sodshocktube}), it is clear that the PPM scheme produces by
far the cleanest solution with all wave families crisply reproduced
(in particular, the contact discontinuity and shock).  \zeus\ and
MUSCL produce similar results, with MUSCL doing a slightly better job
on the rarefaction fan.  In Figure \ref{fig.sodshocktube}, we also
show the integrated absolute deviation from the exact solution,
$||E_1|| = \sum_i \Delta x_i |F(x_i)-{\rm Exact}(x_i)|$.  These
numbers confirm the qualitative differences noted previously.
